---
title: "Practical Machine Learning Course Project"
output:
  md_document
---

## Summary

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, the goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants and  predict the manner in which they did the exercise. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.

## Loading necessary libraries

```{r, message = FALSE}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
library(knitr)
library(gbm)
library(survival)
library(splines)
library(parallel)
library(plyr)
```

## Loading Data

Loading the working data and the 20 test case.

```{r}
if (!file.exists("pml-training.csv")) {
  download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", 
                destfile = "pml-training.csv")
}
if (!file.exists("pml-testing.csv")) {
  download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", 
                destfile = "pml-testing.csv")
}

training <- read.csv("pml-training.csv", sep = ",", na.strings = c("NA","#DIV/0!",""))
testing <- read.csv("pml-testing.csv", sep = ",", na.strings = c("NA","#DIV/0!",""))
```

## Cleaning and Preparing Data

Splitting the training and test data

```{r}
inTrain <- createDataPartition(training$classe, p = 0.6, list = FALSE)
mytraining <- training[inTrain, ]
mytesting <- training[-inTrain, ]
```

Removing variables with near zero variance

```{r}
nzv <- nearZeroVar(mytraining, saveMetrics = TRUE)
mytraining <- mytraining[, nzv$nzv == FALSE]

nzv <- nearZeroVar(mytesting, saveMetrics = TRUE)
mytesting <- mytesting[, nzv$nzv == FALSE]

mytraining <- mytraining[c(-1)]
testing <- testing[-1,]
```

Removing variables that are almost NA's.

```{r}
trainingtemp <- mytraining
for(i in 1:length(mytraining)) {
    if( sum( is.na( mytraining[, i] ) ) /nrow(mytraining) >= .7) {
        for(j in 1:length(trainingtemp)) {
            if( length( grep(names(mytraining[i]), names(trainingtemp)[j]) ) == 1)  {
                trainingtemp <- trainingtemp[ , -j]
            }   
        } 
    }
}

mytraining <- trainingtemp
rm(trainingtemp)

header1 <- colnames(mytraining)
header2 <- colnames(mytraining[, -58])
mytesting <- mytesting[header1]
testing <- testing[header2]   
```

Coercing testing data into the same type.

```{r}
for (i in 1:length(testing) ) {
    for(j in 1:length(mytraining)) {
        if( length( grep(names(mytraining[i]), names(testing)[j]) ) == 1)  {
            class(testing[j]) <- class(mytraining[i])
        }      
    }      
}

# To get the same class between testing and myTraining
testing <- rbind(mytraining[2, -58] , testing)
testing <- testing[-1,]
```
## Prediction with Decision Trees

```{r}
set.seed(74810)
modFitA1 <- rpart(classe ~ ., data = mytraining, method = "class")
fancyRpartPlot(modFitA1)

predictionsA1 <- predict(modFitA1, mytesting, type = "class")
cmtree <- confusionMatrix(predictionsA1, mytesting$classe)
cmtree
```

## Prediction with Random Forests

```{r}
set.seed(74810)
modFitB1 <- randomForest(classe ~ ., data = mytraining)

predictionB1 <- predict(modFitB1, mytesting, type = "class")
cmrf <- confusionMatrix(predictionB1, mytesting$classe)
cmrf
plot(modFitB1)
```

## Prediction with Generalized Boosted Regression

```{r, message = FALSE}
set.seed(74810)
fitControl <- trainControl(method = "repeatedcv", number = 5, repeats = 1)

gbmFit1 <- train(classe ~ ., data = mytraining, method = "gbm", trControl = fitControl, verbose = FALSE)
gbmFinMod1 <- gbmFit1$finalModel

gbmPredTest <- predict(gbmFit1, newdata = mytesting)
gbmAccuracyTest <- confusionMatrix(gbmPredTest, mytesting$classe)
gbmAccuracyTest
```

## Predicting Results on the Test Data

Ramdom Forests gave an Accuracy of 99.71% so this model was used to predict the 20 cases.

```{r}
predictionB2 <- predict(modFitB1, testing, type = "class")
predictionB2
```